---
title: "Spatial Data Basics"
subtitle: "Prepared for PAP New Jersey Trainings"
date: "February 10, 2022"
author: Kelly Pierce, Texas Advanced Computing Center
output:
  pdf_document:
    toc: yes
    toc_depth: '6'
  html_document:
    toc: yes
    theme: united
    toc_depth: 6
    toc_float: yes
---


```{r knit, include=FALSE}
library(knitr)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, root.dir = '~/data_trainings/PAP-TACC-2022/data/')
```

# Coordinate reference systems (CRS) and projected reference systems

We use two reference systems, depending on the task:

- WGS84 coordinate reference system is *unprojected*; it references points on an ellipsoid surface. Paradoxically, we use this reference system for drawing maps because the functions we use for map-making understand this reference system and know how to project it to a planar surface. Coordinates in this reference system are measured in decimal degrees or degrees:minutes:seconds.

- The New Jersey State Planar *projected* reference system is locally tailored for the state of New Jersey for accurate spatial representations, measured in meters, and can be used for euclidean distance calculations that would otherwise be distorted if attempted on an ellipsoid reference system such as WGS84. However, mapping functions generally do not know how to interpret this reference system.

The ArcGIS blog has a nice explanation of geographic coordinate systems and projected coordinate systems: https://www.esri.com/arcgis-blog/products/arcgis-pro/mapping/gcs_vs_pcs/ if you'd like further reading. 

These coordinate reference systems can be referred to by standardized codes. We'll define two constants referring to those codes here, and we'll refer to the coordinate references by these variable names later on.

```{r geometry-constants, include=TRUE, echo=TRUE}
WGS84 = 4326
NJ_PLANAR = 'ESRI:102311'
```

# Example: New Jersey Bus Stops

We'll use New Jersey bus stop locations to explore some of the analyses you can perform with spatial data. Navigate to New Jersey Geographic Information Network (https://njogis-newjersey.opendata.arcgis.com/datasets/NJTRANSIT::bus-stops-of-nj-transit/explore) and download the shapefile dataset by clikcing on the download icon in the right-hand panel and selecting the "Shapefile" option. This will download the data as a `zip` archive. Unpack the `zip` archive and make note of the directory path. As you work through these exercises on your own, you can either move the shapefile directory to your working directory and load with a relative file path, or load the data by specifying its full file path.

## Load geographic data from CSV and convert to simple features.

We will use the built-in `read.csv()` function to load our data. This function needs only the path to a valid CSV file (a file where columns are separated by "," and rows are separated by newlines) to load our data. Then we will use the `st_as_sf()` function from the library `sf` to convert the data to a simple features object which can represent geographic data. Other functions from the library `sf` will help us explore this dataset. Read more about the `sf` library [here](https://r-spatial.github.io/sf/articles/sf1.html).

```{r}
bus_csv <- read.csv('data/Bus_Stops_of_NJ_Transit.csv')
head(bus_csv)
```

After inspecting the data, it appears the latitude and longitude data are in columns `DLAT_GIS` and `DLONG_GIS`. We have to tell `st_as_sf()` which CRS to use; unfortunately, CSVs do not have CRS metadata so we'll assume the common WGS84 CRS was used.

```{r}
library(sf)
bus_sf <- bus_csv %>% st_as_sf(.,coords=c("DLONG_GIS","DLAT_GIS"), crs=WGS84)
head(bus_sf)
```

We see now that a `geometry` column has been added to our dataframe, with coordinate `POINT` values. We could save this file using the `st_write()` function (though we will skip that for this exercise). We will use the actual shapefile for the remainder of our work, so we remove the `bus_sf` variable to avoid confusion:

```{r}
rm(bus_sf)
```

## Load a shapefile from file

We will use the `st_read()` function from the library `sf`, which we loaded previously. We're also going to load the `tidyverse` library, which has additional useful methods for data exploration.

```{r}
library(tidyverse)
bus <- st_read('data/Bus_Stops_of_NJ_Transit.shp')
```

## Identify the coordinate reference system and geometry type

This dataset is loaded as a "simple feature collection", but you can treat it almost the same as a dataframe or `tbl_df`. This dataset does have two additional features that make it better suited for spatial data analysis than a regular dataframe:

Shapefiles contain metadata about the coordinate reference system (CRS) that are loaded with the raw data are loaded. You may have noticed some of this metadata printed when we loaded the data; we can als inspect the CRS with the `st_crs()` function:

```{r}
st_crs(bus)
```

Second, simple feature collections contain a special `geometry` column that contains spatial data. If we look at the first element of the `bus` dataset `geometry` column, we see that this column contains `MULTIPOINT` data. It's unclear why `st_read()` specified this as `MULTIPOINT` data, but this won't matter for the work we're doing. Still, if we wanted to convert the `geometry` column to `POINT` data we could do so with `st_cast(bus$geometry, "POINT")`.

```{r}
bus$geometry[1]
```

All values in the `geometry` column must be of the same type, so even though this first entry is a single point there are likely other bus stops in the data set that have multiple points. We'll work to understand the contents of the `geometry` column later in this example.

## Change the coordinate reference system

At this point you may have noticed that the values in the `geometry` column don't look like regular latitude and longitude values. Indeed, the CRS for the bus stop data is indicated as "WGS 84 / Pseudo-Mercator". This is a planar projection of the data, but not the one we want. We can use the `st_transform()` function to do the geometric calculations to translate between coordinate reference systems.

We'll make a new data object in the WGS84 CRS:

```{r}
bus_wgs84 <- st_transform(bus, crs=WGS84)
```

We'll also make a new data object in the NJ Planar projection:

```{r}
bus_nj <- st_transform(bus, crs=NJ_PLANAR)
```

We can double check that our transformations changed the data by looking at the `head()` of our new datasets and inspecting the CRS metadata:

```{r}
head(bus_wgs84)
```

```{r}
head(bus_nj)
```

## Map the bus stops

### Basic point map

We'll use the library `ggplot2` to build our map. `ggplot2` is part of the `tidyverse` set of packages, and uses layers to construct images. Our first layer is a blank plotting space, which we create by calling the `ggplot()` function without any arguments. We then add the actual data we want to visualize in another layer with the `geom_sf` function, which understands how to interpret the `geometry` column in `bus_wgs84`.

```{r message=FALSE, warning=FALSE}
nj_bus_map <- ggplot() +
  geom_sf(data=bus_wgs84, aes(color=STOP_TYPE), inherit.aes = FALSE)
nj_bus_map
```

### Add a basemap

The bus stop points alone are hard to interpret, so we can add information to our map by adding a basemap -- a background layer showing streets and landmarks to help contextualize the bus stop data. Basemap tiles can be downloaded from Stamen Maps.

#### Get a bounding box for the basemap

We have to specify a bounding box, the coordinates for a box that contains the state of New Jersey, in order to download the basemap tiles. The easiest way to get a vaild bounding box is to use the `st_bbox()` function on a polygon to extract the bounding box coordinates.

We will load polygons for the state of New Jersey and its county boundaries directly from the ArcGIS Open Data by downloading a `geojson` file from a URL. `geojson` files are specially formatted text files that can be interpreted by the `read_sf()` function. We will transform these data to the WGS84 CRS.

```{r}
nj <- read_sf(
  'https://opendata.arcgis.com/datasets/5f45e1ece6e14ef5866974a7b57d3b95_1.geojson'
  )
nj <- nj %>% st_transform(crs=WGS84)
```

#### Load street data basemap

Now that we have our New Jersey polygon, we can determine its bounding box and use the `get_stamenmap()` function to download a stylized street map background from Stamen Maps.

```{r echo=TRUE, cache=TRUE, results='hide', message=FALSE, warning=FALSE}
nj_bbox <- unname(st_bbox(nj))
nj_base_map <- get_stamenmap(bbox=nj_bbox, maptype="toner", force=TRUE)
```

### Map bus stops on streetmap

Next we'll build a map of bus stops in with the same layer-based approach we used for the basic point map example above. This time we will make our first layer with the basemap by calling the `ggmap()` function from the `ggmap` library and passing our base map variable as an argument. Our next layer will be the bus stop data, and for this example we will color points by the bus stop type (one of the columns in the dataset). The argument `inherit.aes = FALSE` instructs to the function to ignore default plot aesthetics in favor of the aesthetic instructions we have provided.

```{r message=FALSE, warning=FALSE}
library(ggmap)
nj_bus_map <- ggmap(nj_base_map) +
  geom_sf(data=bus_wgs84, aes(color=STOP_TYPE), inherit.aes = FALSE)
nj_bus_map
```

### Population choropleth map

This dataset happens to include a number of measures for each county. Choropleth maps shade polygons by values. We'll take one measure, `POP2010` (the population of each county from the 2010 decennial census) and make a choropleth map. 

```{r}
head(nj)
```

```{r}
nj_map <- ggmap(nj_base_map) + 
  geom_sf(data=nj, aes(fill=POP2010), inherit.aes = FALSE, color = NA, alpha = 0.8)
nj_map
```

## Aggregate bus stops into a grid

Rather than viewing the individual points where bus stops are located, we can aggregate bus stops spatially. By counting the number of stops in different areas, we may see patterns about service level coverage that are hard to pick out when viewing stop locations individually.

We will first create a grid that covers the state of New Jersey, then we will identify the bus stops that fall inside each grid cell.

### Create a five-mile "fishnet" grid

We will need the following elements to build our fishnet grid:
1. The coverage area, in our case the entire state of New Jersey
2. The width of each square in the grid, specified in meters

We will use the `NJ_PLANAR` projection because we want a flat grid. This projection measures spatial distances in meters, and so we specify our fishnet width in meters. We would like to overlay a five-mile grid on the state of New Jersey, so we convert five miles into meters using the constant of 1,609 meters per mile.

```{r}
nj_flat <- st_transform(nj, crs=NJ_PLANAR)
fishnet_width <- 1609 * 5
```

Next we need to adjust our New Jersey shapefile so it contains only the state border, and not the internal county borders. The `st_union()` function accomplishes this. We also use the `st_make_valid()` function to correct any loops or places where the border appears to cross itself.

```{r}
nj_border <- st_make_valid(st_union(nj_flat))
```

Now that we have our coverage area and our fishnet grid size determined, we are ready to use the `st_make_grid()` function to generate our grid. After we make the grid, we will assign each cell a unique identifier (a "net_id"). We will remove grid cells that fall outside the state border, and reshape the list of grid objects into a dataframe that is easier to work with.

```{r}
net <- st_make_grid(
  x=nj_border, cellsize=fishnet_width, what='polygons', square=TRUE, crs=NJ_PLANAR
  )
net_agg <- st_as_sf(net) %>% tibble::rowid_to_column(., "net_id")
net_intersect <- st_intersects(nj_border, net_agg) 
fishnet <- net_agg[unique(unlist(net_intersect)),]
```

We can make a quick visual inspection of our map with `ggmap`:

```{r}
ggmap(nj_base_map) +
  geom_sf(
    data=st_transform(fishnet, WGS84),
    inherit.aes=FALSE, color='black', alpha=0.3
    )
```

### Point-in-polygon join

With our fishnet created, we're ready to join our bus stop data. We use the `st_join()` function to identify the bus stops in each fishnet cell. This join preserves only the bus stop geometry, but we're really interested in the grid square shapes. So we drop the geometry column -- we will add it back later with another join.

```{r}
bus_net <- st_join(bus_nj, fishnet, join=st_within) %>%
    st_drop_geometry() 
head(bus_net)
```

Now that we have bus stops associated with `net_id` values, we can count the number of bus stops in each cell by grouping rows by the `net_id` value. To make our analysis a bit more interesting, we will keep the `STOP_TYPE` data by using that as a grouping variable as well. This allows us to look at bus stop counts by stop type to see how service type varies spatially. The `summarise(count=n())` operation is what performs this count aggergation by counting the rows in each group.

```{r}
bus_net_summary <- bus_net %>% 
  group_by(net_id, STOP_TYPE) %>% 
  summarise(count=n())
```

Earlier we dropped the `geometry` column, but now we're ready to add it back. We specifically want to add the fishnet `geometry` (not the bus stop point locations), so we'll join our `bus_net_summary` data to our original `fishnet` on the column `net_id`. We have to wrap that join in the `st_as_sf()` function to ensure the resulting dataframe is an `sf` object; we also specify that we would like to use the CRS already present in the `fishnet` data.

```{r}
bus_net_final <- st_as_sf(
  left_join(
    fishnet, 
    bus_net_summary, 
    by='net_id'
    ), 
  crs=st_crs(fishnet))
```

## Map gridded bus stop counts by type

Now we're ready to map our data. We will use the `nj_base_map` as the first layer of our map, and we'll use the fishnet grids as the second layer. The `facet_wrap()` function allows us to easily make separate panels to display each bus stop type individually. We'll spread our facets over two rows for easier viewing:

```{r warning=FALSE, message=FALSE, fig.height=11, fig.fullwidth=TRUE}
bus_type_map <- ggmap(nj_base_map) + 
  geom_sf(
    data=st_transform(bus_net_final, crs=WGS84),
    aes(fill=count),
    inherit.aes = FALSE, alpha=0.8
    ) +
  facet_wrap(facets='STOP_TYPE', nrow=3)
bus_type_map
```

# Exercise: New Jersey Hospital Locations

1. Download "Hospitals in NJ" from the NJGIN Open Data portal: https://njogis-newjersey.opendata.arcgis.com/datasets/hospitals-in-nj/explore?location=40.145600%2C-74.726050%2C8.85&showTable=true

2. Load and inspect the data.
    a. What is the CRS used?
    b. How many columns are in the data set?
    c. How many rows?
    d. What are the data types as understood by R? Do those data types match your understanding of the data?
  
3. Make a faceted histogram of beds by region. Adjust the histogram bin size as necessary.

4. Assign each hospital to a grid cell in the 5-mile grid.
    a. Plot the corresponding choropleth map of hospitals per grid cell.
    b. Plot a histogram of the hospital per grid cell count.
    
5. Make a county-level fishnet.
    a. Use the `filter()` method to subset a single county's data from the hospital dataset.
    b. Make a new fishnet grid for that county only, this time with half-mile resolution.
    c. Aggregate that county's hospital data into the half-mile fishnet grid.
    b. Make the corresponding choropleth map and histogram for this county only.

